% Chapter 3

\chapter{Methodology}\label{Chapter3}

%----------------------------------------------------------------------------------------

\section{Data Description}
Data originates from a clinical neuropsychology database managed by a rehabilitation center. Raw data includes 22,075 records from 8,739 unique individuals with repeated follow-up assessments. Each assessment consists of 31 features relating to demographics/administrative details/injury specifics/neuropsychological results across six cognitive domains: Orientation/Attention/Visuospatial Perception/Language/Memory/Executive Function. Following Tier-1 filtering (described in Section~\ref{sec:filtering}), the analyzed dataset contains 17{,}406 assessments from 7{,}285 subjects.

The primary unit of analysis is the assessment rather than the patient. Repeated follow-up assessments are retained as separate rows in the clustering, imputation-comparison, and assessment-level bootstrap analyses; they are not first aggregated within the 7{,}285 patients. This design preserves clinically observed change over time, but it does not account for within-patient dependence and may make stability estimates appear more precise than estimates obtained from a patient-grouped resampling design. The multinomial sensitivity analysis in Section~\ref{sec:h3_results} is the principal exception because its training, test, and bootstrap samples are grouped by patient.



Given that these data originate from routine clinical practice rather than a specifically constructed research protocol, test selection was influenced by both the neuropsychologist conducting testing/referral questions/patient engagement capabilities. Therefore, missingness in the dataset is structured. Certain measures appear in almost every assessment; however, some are only available in response to certain clinical circumstances. Unlike missingness originating from research protocols that may control missingness through design; missingness in this dataset varies greatly; ranging from approximately 5\% for screening tools to over 90\% for specialty assessments.



Additionally, measures fall along varying scales (standardized z-scores/scaled scores/raw timed measures). During domain aggregation (Section~\ref{sec:domain_scores}), these scales were unified. In Table~\ref{tab:missing_descriptive}, I display the missingness distribution for the fourteen candidate variables. Only 53.1\% (9{,}245 assessments) of the 17{,}406 records contained complete data across all fourteen variables; if I applied listwise deletion, 46.9\% (8{,}161 assessments) of the sample could potentially be lost because this portion of the dataset could not be analyzed. The imputation strategy (Section~\ref{sec:imputation}) is intended to avoid losing clinical information associated with this portion of the sample.



All data have undergone deidentification according to established standards for protecting patient privacy.

\section{Tier 1 Filtering}\label{sec:filtering}
Since many variables in the dataset exhibit large amounts of missingness, I performed filtering before imputation/clustering. In general, the primary analysis employs a fifty-percent filter. Any variable that exhibits more than fifty percent missingness is eliminated; any record exhibiting more than fifty percent missingness among remaining variables is eliminated.



This choice represents a compromise between maximizing data retention versus minimizing the amount of missingness placed on imputation models. While employing a seventy-percent filter would retain significantly more records/variables; this would increase the likelihood that artificial structure exists in the imputed data. Conversely, employing a thirty-percent filter would likely produce a clean matrix; however, it would eliminate substantially more clinically relevant records while reducing analytical capability.



I perform sensitivity analyses utilizing filters between thirty percent and seventy percent in Section~\ref{sec:robustness}.



Following application of this filter and elimination of assessments without documented cognitive testing, the dataset contains fourteen neuropsychological variables; seventeen thousand four hundred-six assessments from seven thousand two-hundred-eighty-five patients.

\begin{table}[ht]
\caption{Eligible neuropsychological variables after Tier~1 filtering, organised by cognitive domain.}
\label{tab:eligible_vars}
\centering
\small
\begin{tabular}{l p{0.55\textwidth} c}
\toprule
\textbf{Cognitive Domain} & \textbf{Variables} & \textbf{Count} \\
\midrule
Orientation        & \texttt{OPERSONA}, \texttt{OESPAI}, \texttt{OTEMPS} & 3 \\
Attention          & \texttt{ASPAN}, \texttt{ATMTA} & 2 \\
Visuoperception    & \texttt{VPIMATGES} & 1 \\
Language           & \texttt{LREPETICIOTB}, \texttt{LDENOMINACIOTB}, \texttt{LCOMPRENSIOTB} & 3 \\
Memory             & \texttt{MDIGITS}, \texttt{MRAVLT075}, \texttt{MRAVLT015}, \texttt{MRAVLT015R} & 4 \\
Executive Function & \texttt{FEPMR} & 1 \\
\midrule
\textbf{Total}     & & \textbf{14} \\
\bottomrule
\end{tabular}
\end{table}

\section{Imputation Methods}\label{sec:imputation}

The next part of this section presents the ten different types of multiple imputation methods, grouped into the five categories outlined previously. For clustering performance, both silhouette score and number of clusters \(k\), I chose the optimal combination of UMAP+HDBSCAN using the domain-level features. The silhouette value calculated here is based only on the high-density region clustered by HDBSCAN before performing the \(k\)-nearest-neighbour noise-assignment process. Consequently, the silhouette value is somewhat higher than reported in the literature for fully covered assessments (see Section~\ref{sec:h1_results}); for example, MICE achieved 0.598 in the high-density area before reassignment and 0.526 when assessed across all assigned cases.

The two deep learning methods required additional training choices, so their technical details are specified here. For the denoising autoencoder (DAE), deliberately corrupted inputs were generated by masking the observed values. The corruption mask was sampled as follows:

\begin{equation}\label{eq:dae_corrupt}
\mathbf{m}_i\sim
\begin{cases}
P_{\mathrm{miss}} & \text{with probability } 0.5,\\[2pt]
\operatorname{Bernoulli}(1-\rho)^{\otimes p} & \text{with probability } 0.5,
\end{cases}
\end{equation}

where \(P_{\mathrm{miss}}\) denotes the empirical distribution of missingness patterns and \(\rho\) denotes the uniform corruption rate. This scheme allows the DAE to learn from random corruption as well as from missingness structures that resemble those observed in the clinical data.

For the variational autoencoder (VAE), cyclical annealing was used for the \(\beta\) term in the evidence lower bound. The annealing schedule was:

\begin{equation}\label{eq:beta_cyclical}
\beta(t)=\beta_{\max}\,\frac{t \bmod T}{T},\qquad T=40,\ \beta_{\max}=0.1,
\end{equation}

The deep-learning imputations were evaluated by holding out a subset \(H\) of observed entries and comparing the imputed values with the original values. Root mean square error and mean absolute error were computed as:

\begin{equation}\label{eq:rmse}
\mathrm{RMSE}=\sqrt{\frac{1}{|H|}\sum_{(i,j)\in H}\bigl(\hat{x}_{ij}-x_{ij}\bigr)^2},\qquad
\mathrm{MAE}=\frac{1}{|H|}\sum_{(i,j)\in H}\bigl|\hat{x}_{ij}-x_{ij}\bigr|.
\end{equation}

\begin{table}[htbp]
\caption{A summary of the ten imputation methods broken down by the five-category taxonomy. The clustering performance columns report the optimal UMAP+HDBSCAN solution on domain-level features, using the high-density HDBSCAN core before \(k\)-nearest-neighbour noise assignment.}
\label{tab:method_summary}
\centering
\scriptsize
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{@{} p{1.8cm} l c c c p{3.2cm} p{3cm} @{}}
\toprule
\textbf{Category} & \textbf{Method} & \textbf{Year} & \textbf{$k$} & \textbf{Sil.} & \textbf{Advantages} & \textbf{Limitations} \\
\midrule
Conventional & Mean       & ---  & 4 & 0.604 & Simplest; fast; universal baseline & Collapses variance \& covariance \\
Statistical  & EM         & 1977 & 4 & 0.587 & Principled MLE; preserves covariance & Assumes multivariate normality \\
             & MICE       & 2011 & 3 & 0.598 & Flexible; preserves distribution & Requires model per variable \\
\midrule
Machine      & KNN        & 2001 & 4 & 0.576 & Non-parametric; captures local structure & Sensitive to $k$; slow for large $n$ \\
Learning     & MissForest & 2012 & 3 & 0.606 & Non-linear; handles mixed types & Computationally expensive \\
\midrule
Hybrid       & PMM        & ---  & 4 & 0.498 & Preserves marginal distribution & Requires adequate donor pool \\
\midrule
Matrix       & SoftImpute & 2010 & 4 & 0.571 & Efficient for sparse matrices & Assumes low-rank structure \\
Completion   & NMF        & 1999 & 3 & 0.546 & Non-negativity natural for scores & Sensitive to rank choice \\
\midrule
Deep         & DAE        & 2018 & 5 & 0.510 & Non-linear; multiple imputations & Requires hyperparameter tuning \\
Learning     & VAE        & 2014 & 4 & 0.603 & Probabilistic; principled uncertainty & Complex training; $\beta$-tuning \\
\bottomrule
\end{tabular}

\vspace{0.5em}
\raggedright
\scriptsize
\textit{References:} Little and Rubin \parencite{little2019statistical}; PMM and MICE \parencite{vanbuuren2011mice}; EM \parencite{dempster1977maximum}; KNN \parencite{troyanskaya2001missing}; MissForest \parencite{stekhoven2012missforest}; SoftImpute \parencite{mazumder2010spectral}; NMF \parencite{lee1999learning}; DAE \parencite{gondara2018mida}; VAE \parencite{kingma2014auto, mccoy2018variational}.
\end{table}

\section{Calculating Scores for Each Cognitive Domain}\label{sec:domain_scores}
A set of cognitive-domain-level features was created to represent aggregate versions of the variables related to each cognitive domain, while retaining the same 14 individual neuropsychological variables for comparison. Since many of the original measures are recorded on different scales (e.g., standardised scores versus scaled scores), direct aggregation may create scale-related problems unless scale alignment is performed first. To reduce potential problems from extreme imputations caused by outliers or sentinel codes, three processing steps were applied to each variable before aggregation:

\begin{enumerate}
\item Winsorizing each variable over the 1st to 99th percentile range.
\item Reversing the sign of all timed test results, so that positive values indicate better performance.
\item Standardising all variables to a mean of 0 and variance of 1.
\end{enumerate}

Administration-frequency-weighted averages of variables representing each domain were then used as domain scores. Administration-frequency weights (\(w_j = 1 - r_j\), where \(r_j\) represents the missingness rate) were assigned to each variable \(j\) and normalised within domains. This weighting gives less influence to sparsely administered measures, whose composite contribution would otherwise depend more heavily on imputed values, and gives more influence to measures observed more often in routine practice. The weights should not be interpreted as psychometric reliability estimates: they do not quantify internal consistency, test-retest reliability, or measurement error. By reducing dimensionality from 14 individual variables to six domain-level scores, the number of variables available for distance-based algorithms is reduced, improving feature-space geometry. Furthermore, creating domain-level profiles facilitates clinical conceptualisation of cognitive deficit in relation to cognitive domains rather than individual variables. Some detail is sacrificed. For example, two patients could obtain equal average memory scores while differing in immediate versus delayed recall. Whether there is sufficient rationale for sacrificing this detail in exchange for reduced redundancy and improved interpretability is assessed in Hypothesis H4.

\section{UMAP + HDBSCAN Pipeline}\label{sec:pipeline}
Figure~\ref{fig:pipeline_diagram} shows the end-to-end process from imputed data to final cluster assignment. Prior to performing dimensional reduction using UMAP/HDBSCAN, I apply robust scaling using the median and interquartile range to reduce influence of floor and ceiling effects common in neuropsychological assessment \parencite{bellio2020analyzing,pinsky2023mad}. Floor effects occur when severely impaired patients accumulate at minimum values for an assessment tool \parencite{pinsky2023mad}, whereas ceiling effects occur when patients with no impairments accumulate at maximum values \parencite{pinsky2023mad}. Scale-free metrics like median and inter-quartile ranges help mitigate these effects. For each of the twenty imputed datasets (ten methods $\times$ two feature representations), I perform UMAP followed by HDBSCAN. Like other machine learning algorithms that rely heavily on initial conditions (in this case, random initialization), UMAP is stochastic. Therefore, a single embedding could capture variability due to either random initialization or geometric properties of the data. Therefore, I repeat each hyperparameter combination ten times and choose the embedding with highest silhouette score (based on initial HDBSCAN clustering).

\begin{center}
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  box/.style={draw=blue!60, line width=0.6pt, rounded corners,
              fill=blue!10, text width=1.9cm, minimum height=1.5cm,
              align=center, font=\footnotesize},
  arr/.style={->, line width=0.9pt, draw=blue!70, >=latex}
]
  \node[box] (imp) at (0,0)    {Imputed data\\(10 methods)};
  \node[box] (std) at (2.6,0)  {Robust scaler\\(median, IQR)};
  \node[box] (umap) at (5.2,0) {UMAP\\9 configs\\$\times$ 10 seeds};
  \node[box] (hdb) at (7.8,0)  {HDBSCAN\\120 configs};
  \node[box] (sel) at (10.4,0) {Best $Q$\\selection};
  \node[box] (knn) at (13.0,0) {KNN noise\\reassignment};
  \draw[arr] (imp) -- (std);
  \draw[arr] (std) -- (umap);
  \draw[arr] (umap) -- (hdb);
  \draw[arr] (hdb) -- (sel);
  \draw[arr] (sel) -- (knn);
\end{tikzpicture}%
}
\captionof{figure}[Pipeline overview]{This is an end-to-end clustering pipeline. Once the data have been imputed, they first pass through a robust scaler. Next, after being scaled, the imputed data pass into a UMAP hyperparameter grid with 10 random seeds for each of nine configurations, and the resulting embeddings that yield the highest silhouette value are retained. HDBSCAN then does the clustering in a 120-configuration sweep, based upon the composite quality measure $Q = \text{silhouette} \times (1 - \text{noise fraction})$, where all boundary noise points are assigned back to their respective cluster(s) via k-nearest neighbour. This entire process is run over each of the 10 possible imputation outcomes.}
\label{fig:pipeline_diagram}
\end{center}

\subsection{UMAP Parameter Grid}
I utilized Euclidean distance as the distance metric and \texttt{min\_dist}$=0.0$ to encourage separation between dense areas in the embedding.

\begin{table}[ht]
\caption{UMAP hyperparameter grid. All combinations of these values were put through their paces, for a total of 9 configurations.}
\label{tab:umap_grid}
\centering
\begin{tabular}{l l}
\toprule
\textbf{Parameter} & \textbf{Values} \\
\midrule
\texttt{n\_neighbors}  & 15, 30, 50 \\
\texttt{min\_dist}     & 0.0 \\
\texttt{n\_components} & 3, 5, 8 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{HDBSCAN Parameter Grid}
Each UMAP embedding produced 120 possible candidate solutions (due to 120 different possible HDBSCAN configurations). To determine which configuration was "best," I employed a composite quality score that punished solutions with high silhouettes by allocating many assignments to noise:

\begin{equation}
Q = \text{silhouette} \times (1 - \text{noise proportion})
\label{eq:quality}
\end{equation}

\begin{table}[ht]
\caption{HDBSCAN hyperparameter grid. All listed value combinations were evaluated, giving me 120 configurations for each UMAP embedding.}
\label{tab:hdbscan_grid}
\centering
\begin{tabular}{l l}
\toprule
\textbf{Parameter} & \textbf{Values} \\
\midrule
\texttt{min\_cluster\_size}          & 500, 1000, 1500, 2000, 3000 \\
\texttt{min\_samples}                & 5, 10, 25, 50 \\
\texttt{cluster\_selection\_method}  & eom, leaf \\
\texttt{cluster\_selection\_epsilon} & 0.0, 0.1, 0.5 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Sweep Procedure}
The following procedures were employed for each imputed dataset. First, I preprocessed each dataset using a robust scaler. Second, I repeated each UMAP hyperparameter combination with ten random seeds, yielding ten embeddings per combination, and retained the embedding with the highest silhouette score. These repetitions assess sensitivity to stochastic initialisation; they are not folds of cross-validation. Third, I evaluated a grid of 120 HDBSCAN parameter configurations and selected the configuration yielding the highest quality score \(Q\). This is a hyperparameter sweep rather than 120-fold cross-validation. Fourth, I re-labelled observations previously labelled as noise. Noise points assigned by HDBSCAN were relocated to the nearest substantive cluster using k-nearest-neighbour similarity (\(k = 10\)) in UMAP space. This process yielded 100\% coverage. Re-labelling was allowed only when the silhouette decreased by less than 10\%, preserving separation between clusters.

\subsection{Worked Example}
Assume a hypothetical patient completes only those neuropsychological assessments related to orientation and digit span but fails to complete those related to Trail Making Test due to physical limitations or the delayed recall portion of the Rey Auditory Verbal Learning Test due to exhaustion. Upon application of Tier 1 filtering, however, the patient remains in the dataset with eligible variables only. The two missing scores are subsequently filled in by MICE based on observed information. The 14 variable-level scores are then subjected to winsorizing, aligned for direction, standardized, and weighted as defined in Section~\ref{sec:domain_scores} to produce six domain-composite scores. The robust scaler is applied to minimize the effect that floor and ceiling effects may have on aggregate measures. The resulting six-dimensional vector is then mapped into the cohort's UMAP representation space. The patient is then classified into one of three cognitive-severity tiers: Above Average, Near Normal, or Globally Impaired. If a patient falls into a sparse area of low density and was classified as noise, KNN reassignment utilizes neighboring assessments in UMAP-space (\(k = 10\)) and assigns them to the most closely located substantively-clustered tier. The final output includes a tier label and confidence score based on how often each imputation method agrees under hypothesis H2.

\section{Hypothesis Definitions}\label{sec:hypotheses}
\noindent \textbf{H1 (Stability and Discreteness of Cognitive Clusters)}. The UMAP + HDBSCAN pipeline is expected to identify at least two stable cognitive clusters for most imputation techniques. The stability criterion is met when the best solution from at least 8 of the 10 methods has an average silhouette $\geq 0.40$ and a pre-reassignment noise proportion $< 0.30$. These numerical thresholds evaluate partition quality and stability in the selected embedding; they do not by themselves establish that the clusters are discrete phenotypes. Discreteness is evaluated separately using profile shape, principal-component structure, separation in the original six-domain space, and sensitivity of the recovered cluster count.



\noindent \textbf{H2 (Robustness Across Imputation Techniques)}. Solutions for the clustering problems should show reasonable to high agreement among the imputation methods when the embedding geometry and partition model are held fixed. The primary discovery pipeline remains UMAP+HDBSCAN. For H2 only, I use a deliberately fixed secondary agreement diagnostic: a \(k\)-means labelling rule is trained on the MICE embedding and then applied unchanged to all other imputations. H2 is confirmed if the median pairwise ARI and NMI for each of the $\binom{10}{2} = 45$ combinations of methods exceeds 0.50. Although there is considerable flexibility in setting this threshold, it is greater than chance and implies substantial overlap in group membership under a common projection and labelling rule. This hypothesis therefore tests robustness of point assignments under a fixed reference structure; it does not require, or claim, that independently refitted UMAP+HDBSCAN solutions recover the identical number of tiers.



\noindent \textbf{H3 (Relationship Between Clinical Unit \& Cluster Membership)}. Cluster membership is expected to be related to clinical unit. For each imputation technique, a contingency table is created and a chi-square test of independence is conducted. H3 is confirmed if the relationship remains statistically significant at \(\alpha = 0.05\) after a Bonferroni adjustment and Cramér's \(V > 0.10\). As a sensitivity check for potential confounds, a Cochran-Mantel-Haenszel (CMH) test is performed, conditioned on an appropriate covariate.



\noindent \textbf{H4 (Clustering at Domain-Level vs. At Variable Level)}. I expect the cognitive features at the domain level to result in a clustering representation that is more interpretable and at least as well separated as the representation based on cognitive features at the variable level. H4 will be confirmed if the domain-level approach performs at least as well as or better than the variable-level approach on at least two of three metrics: (1) higher silhouette scores; (2) lower mean VIFs; and (3) lower condition numbers. Smaller values of VIF and condition number imply less multicollinearity and better numerical conditioning. Because cross-method agreement has already been assessed in H2, comparisons of feature representations are made in H4 exclusively.

\section{Statistical Tests and Validation}\label{sec:stats}

\subsection{Statistical Conventions}
Each statistical evaluation of hypotheses follows typical statistical practice. Each evaluation assesses the data relative to a null hypothesis \(H_0\), representing a "no-effect" baseline. The p-value is the probability of obtaining a test statistic as extreme as that obtained assuming \(H_0\) to be true; smaller p-values represent stronger evidence against \(H_0\). A predetermined significance level of \(\alpha = 0.05\) serves as the threshold for rejecting \(H_0\). For any parameter, a 95\% CI represents the range that would include the population parameter in 95\% of repeated random samples from the same population using the same methodology.



For applications of bootstrapping, observed data are randomly sampled with replacement to generate a simulated approximation of the sampling distribution of a statistic. The 2.5th and 97.5th percentiles of that simulated distribution define the bootstrap CI. A z-score quantifies how many standard deviations away from the mean of a specified reference distribution a given value lies. With respect to the psychological testing instruments employed here, standardized testing permits scores derived from different instruments to be compared.



Cohen defined standards for interpreting the magnitude of effects, specifically for standardized differences, i.e., Cohen's \(d\). Specifically, \(d = 0.20\) is interpreted as a small effect size, \(d = 0.50\) as a medium-sized effect, and \(d = 0.80\) as a large effect size. When applying Cramér's V as an effect size for evaluating chi-squared tests, \(V = 0.10\) is interpreted as a small effect size, \(V = 0.30\) as a medium effect size, and \(V = 0.50\) as a large effect size.

\begin{equation}\label{eq:cohend}
d=\frac{\bar{x}_1-\bar{x}_2}{s_p},\qquad s_p=\sqrt{\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}},
\end{equation}

\subsection{Tests and Validation Procedures}
\noindent \textbf{Bootstrap Stability}. I conduct an assessment-level bootstrap stability analysis for every imputation method using \(B = 50\) resamplings. Assessments are sampled as rows, so repeated observations from the same patient are not grouped during resampling. For each resampling, I refit HDBSCAN on the cached UMAP embeddings, recording the median silhouette (Eq.~\ref{eq:silhouette}) and pre-KNN noise proportions. Pre-KNN noise proportion is equal to the proportion of observations that do not belong to any meaningful cluster:

\begin{equation}\label{eq:noise}
\text{noise}=\frac{1}{n}\bigl|\{\,i:\text{label}(i)=\text{noise}\,\}\bigr|,
\end{equation}

This is also the value used in defining quality function \(Q\) (Eq.~\ref{eq:quality}). According to H1's pre-defined conditions, a method will be deemed successful if its median silhouette is larger than or equal to 0.40 and its pre-KNN noise proportion is less than or equal to 0.30; at least 8 of the 10 methods will need to satisfy both conditions. To evaluate per-cluster stability, I apply Hennig (2007)'s matching algorithm to the pre-KNN labels using the default threshold (i.e., 0.7).



\noindent \textbf{Pairwise Agreement}. In order to remove variability caused by refitting other parts of the pipeline, I employ a shared reference-pipeline strategy. This is a secondary sensitivity analysis, not the clustering procedure used to discover the primary tiers. One StandardScaler, one UMAP model, and one \(k\)-means model with \(k=3\) are trained on MICE-imputed domain scores. The remaining methods are then standardized, projected onto this MICE-trained pipeline, and classified using that same MICE-trained \(k\)-means model. The use of \(k\)-means here is intentional: it supplies a fixed labelling function so that the imputed values are the only element varying across methods. This design tests whether assessments produced by different imputation methods land in similar regions under a fixed embedding and partition; it does not test whether a newly fitted UMAP+HDBSCAN pipeline would reconstruct the same geometry, density hierarchy, or number of tiers from each imputed dataset. It creates the ARI (Eq.~\ref{eq:ari}) and NMI (Eq.~\ref{eq:nmi}) matrices comparing each method to each other shown as heatmaps in Chapter~\ref{Chapter4}. Confidence for each assessment is defined as:

\begin{equation}\label{eq:consensus}
\mathrm{conf}(i)=\max_{\ell}\frac{1}{M}\sum_{m=1}^{M}\mathbf{1}\!\left[\text{label}_m(i)=\ell\right],
\end{equation}

High-confidence assessments are those whose assessments were placed into the mode cluster by at least \(M/10 = 0.8 \times 10 = 8\) methods.



\noindent \textbf{Chi-Square Test, Cramér's V, CMH Test}. Contingency tables arise from cross-classifying two categorical variables - e.g., cluster assignment and clinical unit - giving joint counts. Pearson's Chi-Square Test evaluates observed counts \(O_{ij}\) against expected counts \(E_{ij}\) under the null hypothesis that the two variables are independent:

\begin{equation}\label{eq:chi2}
\chi^2=\sum_{i=1}^{r}\sum_{j=1}^{c}\frac{(O_{ij}-E_{ij})^2}{E_{ij}},\qquad E_{ij}=\frac{O_{i\cdot}\,O_{\cdot j}}{n},
\end{equation}

with \(O_{i\cdot}\) and \(O_{\cdot j}\) denoting row sums and column sums, respectively. The test statistic is referenced against a Chi-Square distribution with \((r - 1)(c - 1)\) degrees of freedom to yield the p-value. Cramér's V is used to render association statistics scale-free for sample-size purposes, since it takes on values between zero and one:

\begin{equation}\label{eq:cramersv}
V=\sqrt{\frac{\chi^2}{n\,(k-1)}},\qquad k=\min(r,c),
\end{equation}

The Cochran-Mantel-Haenszel Test extends Pearson's Chi-Square Test for data organized into strata (e.g., diagnosis subgroups). The CMH Test determines whether the association between cluster assignments and clinical units remains significant even when stratifying by an appropriate covariate:

\begin{equation}\label{eq:cmh}
\chi^2_{\mathrm{CMH}}=\frac{\bigl(\sum_{h}(a_h-\mathbb{E}[a_h])\bigr)^2}{\sum_{h}\operatorname{Var}(a_h)}.
\end{equation}

\noindent \textbf{Patient-Grouped Multinomial Sensitivity Analysis}. Because association statistics do not establish predictive performance, I also compared two regularised multinomial logistic models for detailed clinical-unit assignment. The baseline model used the broad diagnostic grouping supplied by the clinical-unit lookup; the augmented model added cognitive-severity tier. Units represented by fewer than 50 unique patients were combined into an ``other units'' category, leaving 17 outcome classes. Five-fold stratified group cross-validation kept every assessment from the same patient in a single fold. Held-out multinomial log loss was the primary performance measure, with accuracy and macro-F1 reported descriptively. The difference in out-of-fold log loss was bootstrapped 1{,}000 times by patient. Because the broad diagnostic grouping is derived from the same lookup as detailed clinical unit, this model is a sensitivity analysis of incremental classification performance rather than independent external validation.



\noindent \textbf{Variance Inflation Factor}. To quantify multicollinearity in regression models, VIF is used. For any variable \(j\), it is defined as:

\begin{equation}\label{eq:vif}
\mathrm{VIF}_j=\frac{1}{1-R_j^2},
\end{equation}

where \(R_j^2\) denotes the coefficient of determination when regressing \(j\) on all other predictors. Any value exceeding 10 constitutes grounds for concern, and influences interpretation both at the variable level and in subsequent analyses at the domain level.

\subsection{Implementation}

\begin{sloppypar}
All analyses were implemented in Python (version 3.13).
Imputation used
\texttt{scikit-learn} (KNN, NMF),
\texttt{fancyimpute} (SoftImpute, EM),
\texttt{miceforest} (MICE, PMM), and
\texttt{missforest} (MissForest).
DAE and VAE used \texttt{PyTorch}.
UMAP used \texttt{umap-learn};
HDBSCAN used the \texttt{hdbscan} library.
Statistical tests used \texttt{scipy.stats} and \texttt{statsmodels}.
Validation metrics used \texttt{scikit-learn}.
The full source code, notebooks, and dashboard are available at
\url{https://github.com/zkunos/thesis_project}.
\end{sloppypar}

%----------------------------------------------------------------------------------------
